A Gentle Introduction and Application of Feature-Based Clustering with Psychological Time Series

### 2023-02-24 latex-results/05FeatureClustering.tex
:memo: clustering wording

For the actual k-means clustering, the analysis seeks to minimize the
total within-cluster variation. We used the Hartigan and Wong algorithm,
which is a widely used approach in k-means clustering of time series
features \citep{hartigan1979}. This algorithm is designed to optimize
the clustering of the feature data into \(k\) groups, where \(k\) is a
pre-defined number of clusters. The algorithm starts by randomly
separating the data points into k clusters and then iteratively updates
the assignment of each point to the nearest cluster center until
convergence. The Hartigan and Wong algorithm specifically calculates the
within-cluster variation (\(W\)) of cluster \(C_i\) as the summed
squared Euclidean distances of the feature \(x\) to its assigned cluster
centroid \(\mu_i\):

@@

By summing the within cluster sum of squares from all \(k\) clusters, we
can then derive the total within-cluster sum of square \(WCSS\):

@@

In our case we used the Elbow and the Silhouette method to determine the
recommended number of \(k\) clusters, which for both methods suggested
two clusters (see Supplemental Material A). We then entered the
participants' PC-scores from the previous step into the k-means
algorithm. To avoid local minima we used 100 random initial centroid
positions. In the final solution the k-means analysis assigned 80
participants to cluster 1 and 76 participants to cluster 2.

>>>>

During the k-means clustering itself, the analysis seeks to minimize the
total within-cluster variation. The analysis is designed to optimize the
clustering of the feature data into \(k\) groups, where \(k\) is a
pre-defined number of clusters. We used the Hartigan and Wong algorithm,
which is a widely used algorithm in k-means clustering
\citep{hartigan1979}. The algorithm starts by randomly separating the
data points into k clusters and then iteratively updates the assignment
of each point to the nearest cluster center until convergence. To do so,
the Hartigan and Wong algorithm specifically calculates the
within-cluster variation (\(W\)) of cluster \(C_i\) as the summed
squared Euclidean distances of the feature \(x\) to the closest cluster
centroid \(\mu_i\):

@@

By summing the within-cluster sum of squares from all \(k\) clusters, we
can then derive the total within-cluster sum of square \(WCSS\):

@@

In our case we used the Elbow and the Silhouette method to determine the
recommended number of \(k\) clusters, which for both methods was two
clusters (see Supplemental Material A). We then entered the
participants' PC-scores from the feature reduction step into the k-means
algorithm. To avoid local minima we used 100 random initial centroid
positions. In the final solution the k-means analysis assigned 80
participants to cluster 1 and 76 participants to cluster 2.

### 2023-02-24 sections/05FeatureClustering.Rmd
:memo: clustering wording

For the actual k-means clustering, the analysis seeks to minimize the total within-cluster variation. We used the Hartigan and Wong algorithm, which is a widely used approach in k-means clustering of time series features \citep{hartigan1979}. This algorithm is designed to optimize the clustering of the feature data into $k$ groups, where $k$ is a pre-defined number of clusters. The algorithm starts by randomly separating the data points into k clusters and then iteratively updates the assignment of each point to the nearest cluster center until convergence. The Hartigan and Wong algorithm specifically calculates the within-cluster variation ($W$) of cluster $C_i$ as the summed squared Euclidean distances of the feature $x$ to its assigned cluster centroid $\mu_i$:

@@

By summing the within cluster sum of squares from all $k$ clusters, we can then derive the total within-cluster sum of square $WCSS$:

@@

In our case we used the Elbow and the Silhouette method to determine the recommended number of $k$ clusters, which for both methods suggested two clusters (see Supplemental Material A). We then entered the participants' PC-scores from the previous step into the k-means algorithm. To avoid local minima we used 100 random initial centroid positions. In the final solution the k-means analysis assigned `r table(kmeansOut$cluster)[1]` participants to cluster 1 and `r table(kmeansOut$cluster)[2]` participants to cluster 2.

>>>>

During the k-means clustering itself, the analysis seeks to minimize the total within-cluster variation. The analysis is designed to optimize the clustering of the feature data into $k$ groups, where $k$ is a pre-defined number of clusters. We used the Hartigan and Wong algorithm, which is a widely used algorithm in k-means clustering \citep{hartigan1979}. The algorithm starts by randomly separating the data points into k clusters and then iteratively updates the assignment of each point to the nearest cluster center until convergence. To do so, the Hartigan and Wong algorithm specifically calculates the within-cluster variation ($W$) of cluster $C_i$ as the summed squared Euclidean distances of the feature $x$ to the closest cluster centroid $\mu_i$:

@@

By summing the within-cluster sum of squares from all $k$ clusters, we can then derive the total within-cluster sum of square $WCSS$:

@@

In our case we used the Elbow and the Silhouette method to determine the recommended number of $k$ clusters, which for both methods was two clusters (see Supplemental Material A). We then entered the participants' PC-scores from the feature reduction step into the k-means algorithm. To avoid local minima we used 100 random initial centroid positions. In the final solution the k-means analysis assigned `r table(kmeansOut$cluster)[1]` participants to cluster 1 and `r table(kmeansOut$cluster)[2]` participants to cluster 2.
